Estimating duration using survival analysis

### 2023-06-11 cis-perfect-testing.tex
Clearer discussion of ghost epsisodes

\todo[inline]{Need to discuss interpretation of the ghosts in the individual model, why this is hard, and the implications for setting a prior here. Not even clear in simulation what it should be!}

@@

Double interval censoring is a special case of censoring, a common and well-studied challenge encountered when dealing with survival data.
An event is \emph{censored} if the time of the event can only be bounded; it is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}).
Left and right censoring can be viewed as special cases of interval censoring where the lower bound is negative infinity or the upper bound is positive infinity respectively.
\emph{Double censoring} means that both the originating and terminating event are censored, and is also a well-studied problem.
Doubly censored data is more challenging to handle because the initiating event time and duration must be jointly modelled~\autocite[and references therein]{liSemiparametric}.
In CIS, we have \emph{double interval censoring} meaning that both the initiating and the terminating events are interval censored; approaches are reviewd by \textcite{sunAnalysis,bogaertsSurvival}.
\todo{double interval censoring refs}.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode. \label{perf-test:fig:double-interval-censor}}
\end{figure}

@@

Few studies consider the combination of both double censoring (of any type) and truncation, especially within the human biostatistical literature, although the general frameworks developed in the 1970s do include this case \autocites{turnbullEmpirical}{dempsterMaximum}.
Perhaps the main use of human studies involving double interval censoring and truncation are those estimating the incubation period of HIV/AIDS (the incubation period in that context is the time from being infected with HIV to developing AIDS symptoms).
\Textcite{sunEmpirical,bacchettiNonparametric} discuss this, however, they use a conditional likelihood approach which conditions on both the truncation (a selection effect) and the time of the infection.
The reason for the latter is unclear, perhaps though the infection time is considered to provide little information on the incubation period~\citePersonalComms{Nick Jewell}; a similar approach is taken by \textcite{shenNonparametric}.
Within CIS, detected infections are more likely to occur in the period where there is frequent testing, and hence conditioning on the infection time ovestates the probability of the infection being detected.
Furthermore, the interpretation and justification of a conditional likelihood approach is unclear in a Bayesian framework.

@@

\section{Modelling the duration}\label{perf-test:sec:model}
In this section, I develop a model accounting for double interval censoring and truncation.
I model only individuals with detected episodes.
This approach is based on long-standing methodology~\autocites{heiseyModelling}{dempsterMaximum}{turnbullEmpirical}.
The conceptual approach here is to form a cohort of individuals for which we observe at least one positive test, and then consider a study which had only enrolled these individuals.
The modelling approach then corrects for this selection bias, by considering \emph{ghost} infections that occur in identical individuals but were truncated.

@@

\subsection{Notation and assumptions}
\todo[inline]{Some of this will possibly be defined elsewhere, such as in the introduction where CIS is introduced}
\begin{itemize}
\item
  Time is discrete: $1, \dots, T$.
\item
  We observe $n_a$ episodes of infection, indexed by $i$.
\item
  The $i$ in which episode $i$ occurs is tested at times
  $t_i = \{ t_{i,1}, \dots, t_{i,m_i} \}$.
\item
  Each observed episode has an unknown start of infection time $b_i$
  and end of infection time $e_i$ (the first and last day that the
  individual would test positive due to this episode respectively).
  These are realisations of the random variables $B_i$ and $E_i$
  respectively.
\item
  The duration of the infection is the number of days for which an
  individual tests positive, $D_i = E_i - B_i + 1$.
\item
  We assume that, for all episodes $i$, $D_i$ is iid and independent
  of the time of the infection. Define the survival function
  $\prob(D_i \geq t \mid B_i = b, \theta) = \prob(D_i \geq t \mid \theta) = S_\theta(t)$,
  where $\theta$ are the parameters controlling the survival
  distribution (the discussion in this document is valid regardless of
  the model specified for $S_\theta$ and hence we consider $\theta$
  as an arbitrary vector of parameters).
\item
  The beginning of episode $i$ is known to occur in the interval
  $[l_i^{(b)}, r_i^{(b)}]$, and similarly for the end of the infection
  in $[l_i^{(e)}, r_i^{(e)}]$.
\item
  $y_i(t)$ for $t \in t_i$ is a binary indicator giving the test
  result for individual $i$ at time $t$. Under the perfect testing
  assumption, $y_i(t) = 1$ if and only if $b_i \leq t \leq e_i$.
\item
  Throughout, the convention that lower-case letters are realisations of
  upper-case random variables is used.
\end{itemize}

Episode $i$ can be fully characterised by the pair $(b_i, e_i)$ of
start and end dates of the episode, which belong to the state space
$T \times T$. If the individual in which the episode occurs is tested
between $b_i$ and $e_i$ (inclusive), then we observe the episode,
otherwise it is truncated. For the $i$th episode that we observed, we
consider there is an unknown number, $n_{it}$, where $(b, e)$ are
drawn from the same distribution as those that lead to episode $i$,
and occur in an (imagined) individual identical to $i$, but were
truncated. Episodes are also truncated if there is no negative test
prior to their start, that is, the individual the episode occurred in
had no tests before $b_i$. The truncated episodes are the ghosts of $i$.

>>>>

@@

Double interval censoring is a special case of censoring, a common and well-studied challenge encountered when dealing with survival data.
An event is \emph{censored} if the time of the event can only be bounded; it is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}).
Left and right censoring can be viewed as special cases of interval censoring where the lower bound is negative infinity or the upper bound is positive infinity respectively.
\emph{Double censoring} means that both the originating and terminating event are censored, and is also a well-studied problem.
Doubly censored data is more challenging to handle because the initiating event time and duration must be jointly modelled~\autocite[and references therein]{liSemiparametric}.
In CIS, we have \emph{double interval censoring} meaning that both the initiating and the terminating events are interval censored; approaches are reviewd by \textcite{sunAnalysis,bogaertsSurvival}.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode. \label{perf-test:fig:double-interval-censor}}
\end{figure}

@@

Few studies consider the combination of both double censoring (of any type) and truncation, especially within the human biostatistical literature, although the general frameworks developed in the 1970s do include this case \autocites{turnbullEmpirical}{dempsterMaximum}.
Perhaps the main use of human studies involving both double censoring and truncation are those estimating the incubation period of HIV/AIDS (the incubation period in that context is the time from being infected with HIV to developing AIDS symptoms).
\Textcite{sunEmpirical,bacchettiNonparametric} discuss this, however, they use a conditional likelihood approach which conditions on both the truncation (a selection effect) and the time of the infection.
The reason for the latter is unclear, perhaps though the infection time is considered to provide little information on the incubation period~\citePersonalComms{Nick Jewell}; a similar approach is taken by \textcite{shenNonparametric}.
Within CIS, detected infections are more likely to occur in the period where there is frequent testing, and hence conditioning on the infection time ovestates the probability of the infection being detected.
Furthermore, the interpretation and justification of a conditional likelihood approach is unclear in a Bayesian framework.

@@

\section{Notation and assumptions}
\todo[inline]{To be rewritten with some or all moved to the thesis introduction. Currently just being left here as a reference.}
\begin{itemize}
\item
  Time is discrete: $1, \dots, T$.
\item
  We observe $n_a$ episodes of infection, indexed by $i$.
\item
  The $i$ in which episode $i$ occurs is tested at times
  $t_i = \{ t_{i,1}, \dots, t_{i,m_i} \}$.
\item
  Each observed episode has an unknown start of infection time $b_i$
  and end of infection time $e_i$ (the first and last day that the
  individual would test positive due to this episode respectively).
  These are realisations of the random variables $B_i$ and $E_i$
  respectively.
\item
  The duration of the infection is the number of days for which an
  individual tests positive, $D_i = E_i - B_i + 1$.
\item
  We assume that, for all episodes $i$, $D_i$ is iid and independent
  of the time of the infection. Define the survival function
  $\prob(D_i \geq t \mid B_i = b, \theta) = \prob(D_i \geq t \mid \theta) = S_\theta(t)$,
  where $\theta$ are the parameters controlling the survival
  distribution (the discussion in this document is valid regardless of
  the model specified for $S_\theta$ and hence we consider $\theta$
  as an arbitrary vector of parameters).
\item
  The beginning of episode $i$ is known to occur in the interval
  $[l_i^{(b)}, r_i^{(b)}]$, and similarly for the end of the infection
  in $[l_i^{(e)}, r_i^{(e)}]$.
\item
  $y_i(t)$ for $t \in t_i$ is a binary indicator giving the test
  result for individual $i$ at time $t$. Under the perfect testing
  assumption, $y_i(t) = 1$ if and only if $b_i \leq t \leq e_i$.
\item
  Throughout, the convention that lower-case letters are realisations of
  upper-case random variables is used.
\end{itemize}

\section{Modelling the duration}\label{perf-test:sec:model}

In this section, I develop a model accounting for double interval censoring and truncation.
I model only individuals with detected episodes.
This approach is based on long-standing methodology~\autocites{heiseyModelling}{dempsterMaximum}{turnbullEmpirical}.
The conceptual approach here is to form a cohort of individuals for which we observe at least one positive test, and then consider a study which had only enrolled these individuals.
The modelling approach then corrects for this selection bias, by considering \emph{ghost} infections that occur in identical individuals but were truncated.

@@

Ghost infections are a mathematical convenience to correct for truncated episodes.
Consider that we draw many episodes independently, from a distribution of infection times and durations with all other characteristics identical (most notably the testing schedule and, if the model were to be extended, other covariates of the individual in which the episode occurs).
This probability space is equivalent to any two random variables which can be over any two of infection time, recovery time, and duration.
Conceptually, we will condition on the total number of episodes drawn (which is unknown) corresponding to the detected episode $i$, $n_i$, then consider the probability of observing exactly one episode and this episode occurring in the interval censored ranges observed (being \emph{admissible}), and then finally marginalise over $n_i$ to form the likelihood contribution for this episode.

Unfortunately, there is no clean interpretation of the number of ghost episodes in terms of real-world quantities; in this sense, it is a pure model parameter.
In reality, there is only one individual who is identical to the individual in which the episode occurs in all relevant ways.
Even within the simulation study (see \cref{perf-test:sec:simulation-study}) these are not infections that are simulated, and understanding the ground truth number of ghost infections is not straightforward.

To formalise this, note that all relevant information about episode $i$ can be fully characterised by the (unobserved) pair $(b_i, e_i)$ of start and end dates of the episode, which belong to the state space $T \times T$.
If the individual in which the episode occurs is tested between $b_i$ and $e_i$ (inclusive), then the episode is detected, otherwise it is a ghost episode.
For the $i$th detected episode, assume that there are $n_{it}$ ghost episodes.
Further, we discard episodes if there is no negative test prior to their start, and hence also view any such episodes as ghost episodes.
